---
title: "Application of poisson mash ruv to neutrophils in single cell cytokines data"
author: "Yusha Liu"
header-includes:
date: "2021-4-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Summary
We apply poisson mash ruv to a subset of neutrophils from a single cell cytokines dataset, where the conditions represent different types of cytokine treatments ($R=46$). This dataset includes expression data of $J=7,988$ genes from $9,289$ cells. Genes with fewer than $25$ counts summed over all cells are removed. 
```{r warning=FALSE, message=FALSE}
library(pheatmap)
library(gridExtra)

cols.trt <- readRDS("../applications_sc/B_cells/condition_colors.Rds")
trts <- names(cols.trt)

# swap the color code of IL1b and IL3 for clearer display of data-driven prior covariances
cols.trt2 <- cols.trt
cols.trt2[which(trts=="IL1b")] <- cols.trt[which(trts=="IL3")]
cols.trt2[which(trts=="IL3")] <- cols.trt[which(trts=="IL1b")]
cols.trt <- cols.trt2

```



## Running ED step
To estimate the data-driven prior covariances in the mash prior, we fit poisson mash ruv to a subset of genes that are identified to be differentially expressed across conditions by a conditional multinomial goodness-of-fit test. In the current implementation, the $R$ canonical prior covariances modeling the condition-specific effects are also included in this step with their scaling parameters to be estimated. We expect that in the presence of these canonical prior covariances, we can better estimate the data-driven prior covariances which capture the effect-sharing patterns across conditions that exist in the data.
```{r warning=FALSE, message=FALSE}
## load in the ED results
fit.ed <- readRDS("../applications_sc/Neutrophils/pois_mash_ruv_ed.Rds")

## running time per iteration in ED
fit.ed$runtime/length(fit.ed$ELBO)
                      
## look at the ELBO
plot(fit.ed$ELBO, xlab = "niter", ylab = "elbo", main="fitting ED step")

```



## Running poisson mash ruv
We then fit poisson mash ruv to the entire dataset.
```{r warning=FALSE, message=FALSE}
## load in the poisson mash ruv fit
res <- readRDS("../applications_sc/Neutrophils/pois_mash_ruv_fit.Rds")

## running time
res$runtime

## look at ELBO
plot(res$ELBO, xlab = "niter", ylab = "elbo", main="Fitting poisson mash ruv")

## calculate prior weights
pi.mat <- matrix(res$pi, ncol=length(res$wlist), byrow=TRUE)

rownames(pi.mat) <- c(names(res$Ulist), names(res$ulist))
colnames(pi.mat) <- paste0("w=", round(res$wlist, 3))
pheatmap(pi.mat, cluster_rows=FALSE, cluster_cols=FALSE, fontsize_row = 6, fontsize_col=6, 
         main="Estimated weights of prior covariances in poisson mash ruv")

rowSums(pi.mat, 3)

```



## Estimate of data-driven covariances from ED step
```{r warning=FALSE, message=FALSE}
### Estimate of top eigenvectors of the data-driven covariance initialized by rank-5 PCA
eig.tPCA <- eigen(fit.ed$Ulist[[1]])
pve.tPCA <- eig.tPCA$values/sum(eig.tPCA$values)

for (k in 1:1){
  v <- eig.tPCA$vectors[,k]
  v <- v/v[which.max(abs(v))]
  
  ed.plot <- barplot(v, col = cols.trt, xlab = "", ylim = range(v+0.5*sign(v)),
                     main = paste0(names(fit.ed$Ulist)[1], ": weight ", round(rowSums(pi.mat)[1],3), 
                                   ":\nEigenvector ", k, " (pve = ", round(pve.tPCA[k],3), ")"))
  text(x = ed.plot, y = v + ifelse(v >= 0, 0.1, -0.25),  label = names(cols.trt), pos = 3, cex = 0.75, col = "black", srt=90)
}


### Estimate of top eigenvectors of rank-1 data-driven covariance matrices
for (k in 1:length(fit.ed$ulist)){
  if(rowSums(pi.mat)[k+2] > 1e-1 & sum(fit.ed$ulist[[k]]!=0)>0){
    v <- as.numeric(fit.ed$ulist[[k]])
    v <- v/v[which.max(abs(v))]
    
    ed.plot <- barplot(v, col = cols.trt, xlab = "", ylim = range(v+0.5*sign(v)), 
                       main = paste0(names(fit.ed$ulist)[k], ": weight ", round(rowSums(pi.mat)[k+2],3))) 
    text(x = ed.plot, y = v + ifelse(v >= 0, 0.1, -0.25),  label = names(cols.trt), pos = 3, cex = 0.75, col = "black", srt=90)
  }
}

```



## The list of DE genes
We perform a gene set over-representation analysis of the differentially expressed genes identified by poisson mash ruv at lfsr = 0.01, against all genes in the single cytokines dataset representing the background. These analyses are performed online using [WebGestalt](http://www.webgestalt.org), which can provide non-redundant significantly enriched gene ontology terms. The results can be found [here](../output/Report_Neutrophils.html).
```{r warning=FALSE, message=FALSE}
### load in the poisson mash ruv posterior summary
pois.ruv <- readRDS("../applications_sc/Neutrophils/pois_mash_ruv_posterior.Rds")

### number of DE genes
idx.pois <- which(apply(pois.ruv$lfsr, 1, min) < 0.01)
length(idx.pois)

```



## A few examples (in GO:0002532: production of molecular mediator involved in inflammatory response)
```{r warning=FALSE, message=FALSE}
### load in the plotting function
source("./code/plot_by_gene.R")

### load in the condition-level count data
data <- readRDS("../applications_sc/Neutrophils/data_jr.Rds")
data.jr <- data$data.jr
s <- data$s

plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Alox5ap", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Cd300a", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Fcer1g", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Gpsm3", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Il4ra", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Mapk14", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Myd88", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Snap23", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Tarm1", cols.trt=cols.trt)
plot_by_gene(data=data.jr, s=s, res.pois=pois.ruv, genename="Tlr4", cols.trt=cols.trt)

```
