---
title: "Exploration of over-estimation in poisson mash application to simulated single cell data"
author: "Yusha Liu"
header-includes:
date: "2021-4-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


## Simulation scenario
To simulate single cell data, we started with a subset of the single cell cytokines data. We first randomly permuted the cells with respect to the condition labels to create null data, then used thinning to add condition effects $\beta_{jr} \sim \sum_k \pi_k MVN(0, U_k)$ to a subset of genes, where the prior covariances $U_k$'s are learned from the poisson mash fit to a subset of neutrophils in the original dataset, with the effect size (in terms of $\max_r \mid \beta_{jr}\mid$) scaled to be realistic. We started with the following subsets of the real data, and removed genes with fewer than 25 counts summed over all cells. In each simulated dataset, there are 7,934 genes with 8,823 cells from 25 cytokine treatments. We simulate 20 single cell datasets in total.


## Effect-sharing patterns across conditions for non-null data
```{r warning=FALSE, message=FALSE}
cols.trt <- readRDS("../simulations_sc/setting4/sim_condition_colors.Rds")
Ulist <- readRDS("../simulations_sc/setting4/Ulist_sim.rds")
U <- Ulist$U
u <- Ulist$u


### Top eigenvectors of the full-rank prior covariance 
eig.U <- eigen(U)
pve.U <- eig.U$values/sum(eig.U$values)

for (k in 1:2){
  v <- eig.U$vectors[,k]
  v <- v/v[which.max(abs(v))]
  
  ed.plot <- barplot(v, col = cols.trt, xlab = "", ylim = range(v+0.3*sign(v)),
                     main = paste0("U: weight=", 0.67,  ":\nEigenvector ", k, " (pve = ", round(pve.U[k],3), ")"))

  text(x = ed.plot, y = v + ifelse(v >= 0, 0.1, -0.25),  label = names(cols.trt), pos = 3, cex = 0.8, col = "black", srt=90)
}


### Eigenvector of rank-1 prior covariance 
v <- u/u[which.max(abs(u))]
ed.plot <- barplot(v, col = cols.trt, xlab = "", ylim = range(v+0.3*sign(v)),  main="u: weight=0.33")
text(x = ed.plot, y = v + ifelse(v >= 0, 0.1, -0.25),  label = names(cols.trt), pos = 3, cex = 0.8, col = "black", srt=90)

```



## Summary of effect size over/under-estimation
For poisson mash and poisson mash ruv, we calculate $\frac{E(\beta_{jr}|D)}{\beta_{jr}}$ in the subset of identified unit-condition pairs that are correctly signed, i.e., $\{j,r: \; lfsr_{jr} < 0.05 \; \text{and} \; E(\beta_{jr}|D)\times \beta_{jr} > 0\}$, and report the probability that $\frac{E(\beta_{jr}|D)}{\beta_{jr}} < \frac{1}{c_0}$ and $\frac{E(\beta_{jr}|D)}{\beta_{jr}} > c_0$ for $c_0 = 2, 3, 5, 10$. This analysis is performed on each of the 20 replicate datasets.
```{r warning=FALSE, message=FALSE}
### read in result summary
ratio.pois.mash <- readRDS("../simulations_sc/setting4/sim_signal_v2_pois_mash_no_ruv_estimation_ratio_summary.Rds")
ratio.pois.mash.ruv <- readRDS("../simulations_sc/setting4/sim_signal_v2_pois_mash_ruv_estimation_ratio_summary.Rds")

pois.mash <- matrix(NA, nrow=20, ncol=8)
rownames(pois.mash) <- names(ratio.pois.mash)
colnames(pois.mash) <- c("prob(<0.1)", "prob(<0.2)", "prob(<1/3)", "prob(<0.5)", "prob(>2)", "prob(>3)", "prob(>5)", "prob(>10)")

pois.mash.ruv <- matrix(NA, nrow=20, ncol=8)
rownames(pois.mash.ruv) <- names(ratio.pois.mash.ruv)
colnames(pois.mash.ruv) <- c("prob(<0.1)", "prob(<0.2)", "prob(<1/3)", "prob(<0.5)", "prob(>2)", "prob(>3)", "prob(>5)", "prob(>10)")

for(iter in 1:20){
  ratio <- ratio.pois.mash[[iter]]
  pois.mash[iter,1] <- mean(ratio < 0.1)
  pois.mash[iter,2] <- mean(ratio < 0.2)
  pois.mash[iter,3] <- mean(ratio < 1/3)
  pois.mash[iter,4] <- mean(ratio < 0.5)
  pois.mash[iter,5] <- mean(ratio > 2)
  pois.mash[iter,6] <- mean(ratio > 3)
  pois.mash[iter,7] <- mean(ratio > 5)
  pois.mash[iter,8] <- mean(ratio > 10)
  
  ratio.ruv <- ratio.pois.mash.ruv[[iter]]
  pois.mash.ruv[iter,1] <- mean(ratio.ruv < 0.1)
  pois.mash.ruv[iter,2] <- mean(ratio.ruv < 0.2)
  pois.mash.ruv[iter,3] <- mean(ratio.ruv < 1/3)
  pois.mash.ruv[iter,4] <- mean(ratio.ruv < 0.5)
  pois.mash.ruv[iter,5] <- mean(ratio.ruv > 2)
  pois.mash.ruv[iter,6] <- mean(ratio.ruv > 3)
  pois.mash.ruv[iter,7] <- mean(ratio.ruv > 5)
  pois.mash.ruv[iter,8] <- mean(ratio.ruv > 10)
}

### poisson mash without ruv
round(pois.mash, 3)

### poisson mash ruv
round(pois.mash.ruv, 3)

```